Hybrid recurrent model for regime switching nonstationary time series

### 2021-10-04 iclr2022/hybridmodel_iclr.tex
add hidden regime part in Appendix

Let $\window \in \mathbb{N}$ be the window size, usually this window is equal to the forecast horizon $\window = \lag$. Consider the following  normalized, deseasonalized and rescaled sequence $\rnnwindow^n_T = \{\rnn^{n}_{T-\window+i|T}\}_{1\leqslant i \leqslant w}$, where, for all $1\leqslant n \leqslant N$ and $1\leqslant i \leqslant w$, 
$$
\rnn^{n,T}_{T-w+i|T} = \frac{\ts^n_{T-w+i} -\tspred^{stat,n}_{T-w+i|T-w}}{\meants^n_T}\,,
$$
where $ \meants^n_T = \sum_{i = 1}^{w}  \ts^n_{T-w+i}/w$.
We use this sequence as input of  $\classifier$ and $\predictor$. %Write, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,H\}$, $\err^{n,T}_{T+i|T} = \ts^n_{T+i} - \etspred^{ets,n}_{T+i|T}$.
 Given $\rnnwindow^n_T$, the proposed regime swithing error-corrector reccurent model is given, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,\lag\}$, by :
\begin{align*}
\ts^n_{T+i}  &= \tspred^{stat,n}_{T-w+i|T-w} +  \hat{\hiddenregime}^n_T \cdot \hat{\remainder}^{n}_{T+i} + \varepsilon^n_{T+i}\,, \\
\hat{\hiddenregime}^n_T &= \mathbf{1}_{\classifier(\rnnwindow^n_T) > \threshold}\,, \\
\hat{\remainder}^{n}_{T+i} &= \predictor(\rnnwindow^n_T)_{T+i} \cdot \meants^n_T\,.
\end{align*}

@@

Using a time-series-specific statistical model, the hybrid network achieves striking accuracy on single sequences. Then, the shared recurrent neural network relies on the strength of machine learning methods at leveraging a huge quantities of heterogeneous data.
In addition to the $N$ target time series, we consider $K \times N$ external sequences indexed from $0$ to $T$. For all $1\leqslant n \leqslant N$, $1\leqslant k \leqslant K$ and  $1\leqslant t \leqslant T$, let $\ws^{n,k}_t$ be the value of the $k$-th external sequence at time $t$ linked to the sequence $\fullts^n$. We note $\fullws^n = \{\{\ws_t^{n,k}\}_{1\leqslant t \leqslant T}\}_{1\leqslant k \leqslant K}$ all the values of the family of weak signals linked to the $n$-th sequence. In addition, let $\fullws^n_T = \{\{\ws_{T-w+i}^{n,k}\}_{1\leqslant i \leqslant \window}\}_{1\leqslant k \leqslant K}$ be only the last $\window$ terms of the family. Concatenating $ \rnnwindow^n_T$ and $\fullws^n_T$, we define a new input for our two RNN,   
$$
\fullconcatinput^n_T = \{\concatinput^n_{T-w+i|T}\}_{1\leqslant i \leqslant w}= \{\rnn^n_{T-w+i|T}, \ws^{n,1}_{T-w+i},...,\ws^{n,K}_{T-w+i}\}_{1\leqslant i \leqslant w}\,.
$$
Finally, for all $1\leqslant n \leqslant N$ and for all $i \in \{1,\ldots,\lag\}$ our final model becomes:
%$$
%\tspred^n_{T+i|T} 
%= \etspred^{ets,n}_{T+i|T} +  \rnnmodel(\fullconcatinput^n_T)_{T+i} \times \meants^n_T
%$$
\begin{equation}
\label{eq:full:model}
\ts^n_{T+i}  = \tspred^{stat,n}_{T-w+i|T-w} +  \mathbf{1}_{\classifier(\fullconcatinput^n_T) > \threshold} \cdot \predictor(\fullconcatinput^n_T)_{T+i} \cdot \meants^n_T + \varepsilon^n_{T+i}\,,
\end{equation}
An illustration of our proposed architecture is displayed in Figure~\ref{fig:architecture}.

@@

In our framework, we assume that our sequence are generated with a regime with 2 hidden regime. As this regime, noted $\hiddenregime$, is never observed, we can compute explicitely the following quantity : 
%$p(\ts^n_{T+1:T+h}|\ts^n_{1:T})$

%Based on the Expectation-Maximization (EM) algorithm and the Fisher's identity, we write a new algorithm to estimate jointly $\theta = \{\theta_p,\theta_c\}$, the weights of RNN-p and of RNN-c.

%During our training, we note $\theta^{(t)}$ the value of $\theta$ at an epoch $t$.

%\begin{equation}
%\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h}|\ts^n_{1:T}) = \mathbb{E}\left[\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h},U_T^n|\ts^n_{1:T})\middle|\ts^n_{1:T+h}\right]\,,
%\end{equation}

\textcolor{red}{details to add in appendix.} \textcolor{red}{Assume that $\theta_p$ is the current value for the weights of the RNN-p and the RNN-c and write $\omega_p^n = \mathbb{P}_{\theta_p}(U_T^n = 1|\ts^n_{1:T+h})$. Note that
$$
\omega_p^n \propto \mathbb{P}_{\theta_p}(U_T^n = 1|\ts^n_{1:T})p_{\theta_p}(\ts^n_{T+1:T+h}|U_T^n=1,\ts^n_{1:T})\,.
$$
Then,
\begin{multline*}
\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h}|\ts^n_{1:T}) = \omega_p^n\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h},U_T^n=1|\ts^n_{1:T})\\
+ (1-\omega_p^n)\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h},U_T^n=0|\ts^n_{1:T})\,.
\end{multline*}
The loss function is then given by
\begin{multline*}
\theta\mapsto - \omega_p^n \log p_\theta(\ts^n_{T+1:T+h},U_T^n=1|\ts^n_{1:T})\\
- (1-\omega_p^n) \log p_\theta(\ts^n_{T+1:T+h},U_T^n=0|\ts^n_{1:T})\,,
\end{multline*} 
where no gradient is computed on $\omega_p^n$. In both terms, $\log p_\theta(U_T^n=0|\ts^n_{1:T})$ and $\log p_\theta(U_T^n=0|\ts^n_{1:T})$ are simply given by the output of the FFNN.  Choices for the model.
\begin{itemize}
\item In the case where the  $\varepsilon^n_{T+i}$ are Gaussian and independent, up to an additive constant
$$
\log p_\theta(\ts^n_{T+1:T+h}|U_T^n=0,\ts^n_{1:T}) = -c\sum_{i=1}^{h}(\ts^n_{T+i}  - \tspred^{stat,n}_{T-w+i|T-w})^2
$$
and
\begin{multline*}
\log p_\theta(\ts^n_{T+1:T+h}|U_T^n=1,\ts^n_{1:T})\\
 = -c\sum_{i=1}^{h}(\ts^n_{T+i}  - \tspred^{stat,n}_{T-w+i|T-w} - \meants^n_T\times \rnnmodel(\fullconcatinput^n_T)_{T+i})^2\,.
\end{multline*}
Choice of $c$ ?
\item In the case where the  $\varepsilon^n_{T+i}$ are Laplace and independent, up to an additive constant
$$
\log p_\theta(\ts^n_{T+1:T+h}|U_T^n=0,\ts^n_{1:T}) = -c\sum_{i=1}^{h}|\ts^n_{T+i}  - \tspred^{stat,n}_{T-w+i|T-w}|
$$
and
\begin{multline*}
\log p_\theta(\ts^n_{T+1:T+h}|U_T^n=1,\ts^n_{1:T})\\
 = -c\sum_{i=1}^{h}|\ts^n_{T+i}  - \tspred^{stat,n}_{T-w+i|T-w} - \meants^n_T\times \rnnmodel(\fullconcatinput^n_T)_{T+i}|\,.
\end{multline*}
Choice of $c$ ?
\end{itemize}}

>>>>

Let $\window \in \mathbb{N}$ be the window size, usually this window is equal to the forecast horizon $\window = \lag$. Consider the following  normalized, deseasonalized and rescaled sequence $\rnnwindow^n_T = \{\rnn^{n}_{T-\window+i|T}\}_{1\leqslant i \leqslant w}$, where, for all $1\leqslant n \leqslant N$ and $1\leqslant i \leqslant w$, 
$$
\rnn^{n,T}_{T-w+i|T} = \frac{\ts^n_{T-w+i} -\tspred^{stat,n}_{T-w+i|T}}{\meants^n_T}\,,
$$
where $ \meants^n_T = \sum_{i = 1}^{w}  \ts^n_{T-w+i}/w$.
We use this sequence as input of  $\classifier$ and $\predictor$. %Write, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,H\}$, $\err^{n,T}_{T+i|T} = \ts^n_{T+i} - \etspred^{ets,n}_{T+i|T}$.
 Given $\rnnwindow^n_T$, the proposed regime swithing error-corrector reccurent model is given, for all $1\leqslant n \leqslant N$ and all $i \in \{1,\ldots,\lag\}$, by :
\begin{align*}
\ts^n_{T+i}  &= \tspred^{stat,n}_{T-w+i|T} +  \hat{\hiddenregime}^n_T \cdot \hat{\remainder}^{n}_{T+i} + \varepsilon^n_{T+i}\,, \\
\hat{\hiddenregime}^n_T &= \mathbf{1}_{\classifier(\rnnwindow^n_T) > \threshold}\,, \\
\hat{\remainder}^{n}_{T+i} &= \predictor(\rnnwindow^n_T)_{T+i} \cdot \meants^n_T\,.
\end{align*}

@@

Using a time-series-specific statistical model, the hybrid network achieves striking accuracy on single sequences. Then, the shared recurrent neural network relies on the strength of machine learning methods at leveraging a huge quantities of heterogeneous data.
In addition to the $N$ target time series, we consider $K \times N$ external sequences indexed from $0$ to $T$. For all $1\leqslant n \leqslant N$, $1\leqslant k \leqslant K$ and  $1\leqslant t \leqslant T$, let $\ws^{n,k}_t$ be the value of the $k$-th external sequence at time $t$ linked to the sequence $\fullts^n$. We note $\fullws^n = \{\{\ws_t^{n,k}\}_{1\leqslant t \leqslant T}\}_{1\leqslant k \leqslant K}$ all the values of the family of weak signals linked to the $n$-th sequence. In addition, let $\fullws^n_T = \{\{\ws_{T-w+i}^{n,k}\}_{1\leqslant i \leqslant \window}\}_{1\leqslant k \leqslant K}$ be only the last $\window$ terms of the family. Concatenating $ \rnnwindow^n_T$ and $\fullws^n_T$, we define a new input for our two RNN,   
$$
\fullconcatinput^n_T = \{\concatinput^n_{T-w+i|T}\}_{1\leqslant i \leqslant w}= \{\rnn^n_{T-w+i|T}, \ws^{n,1}_{T-w+i},...,\ws^{n,K}_{T-w+i}\}_{1\leqslant i \leqslant w}\,.
$$
Finally, for all $1\leqslant n \leqslant N$ and for all $i \in \{1,\ldots,\lag\}$ our final model becomes:
%$$
%\tspred^n_{T+i|T} 
%= \etspred^{ets,n}_{T+i|T} +  \rnnmodel(\fullconcatinput^n_T)_{T+i} \times \meants^n_T
%$$
\begin{equation}
\label{eq:full:model}
\ts^n_{T+i}  = \tspred^{stat,n}_{T-w+i|T} +  \mathbf{1}_{\classifier(\fullconcatinput^n_T) > \threshold} \cdot \predictor(\fullconcatinput^n_T)_{T+i} \cdot \meants^n_T + \varepsilon^n_{T+i}\,,
\end{equation}
An illustration of our proposed architecture is displayed in Figure~\ref{fig:architecture}.

@@

In our framework, we assume that our sequence are generated with 2 hidden regime. As this regime, noted $\hiddenregime$, is never observed, we use the Expectation Maximization (EM) algorithm to write our loss function:
We note $\rnnparam^{(t)} = \{\rnnparam_c^{(t)},\rnnparam_p^{(t)}\}$ the weights of $\classifier$ and of $\predictor$ at a fixed epoch $t$.

As the EM algorithm, we want to compute the following value :

\begin{align}
\mathbb{Q}^n(\rnnparam|\rnnparam^{(t)}) &= \mathbb{E}\left[\nabla_\theta \log p_\theta(\ts^n_{T+1:T+h},U_T^n|\ts^n_{1:T})\middle|\ts^n_{1:T+h},\rnnparam^{(t)}\right]\ \\
& = \sum_{U_T^n} p(U_T^n|\ts^n_{1:T+h},\rnnparam^{(t)}) \times \log p_\theta(\ts^n_{T+1:T+h},U_T^n|\ts^n_{1:T}) \\
& = \sum_{U_T^n} p(U_T^n|\ts^n_{1:T+h},\rnnparam^{(t)}) \times \log[p_\theta(U_T^n|\ts^n_{1:T})p_\theta(\ts^n_{T+1:T+h}|U_T^n,\ts^n_{1:T})]
\end{align}

where $ \mathbb{P}_{\theta}(U_T^n = 1|\ts^n_{1:T})$ is parameterized by $\classifier$ and the predictive probability density $p_{\theta}(\ts^n_{T+1:T+h}|U_T^n=1,\ts^n_{1:T})$ is parameterized by $\predictor$ and the statistical model.

Write $\omega_{\rnnparam^{(t)}}^n = \mathbb{P}(U_T^n = 1|\ts^n_{1:T+h},\rnnparam^{(t)})$. We have:

\begin{align*}
\omega_{\rnnparam^{(t)}}^n &= \mathbb{P}_(U_T^n = 1|\ts^n_{1:T+h},\rnnparam^{(t)}) \\
& =  \frac{p(\ts^n_{T+1:T+h},U_T^n = 1|\ts^n_{1:T},\rnnparam^{(t)})}{p_(\ts^n_{T+1:T+h}|\ts^n_{1:T})} \\
& =  \frac{p(\ts^n_{T+1:T+h}|U_T^n = 1,\ts^n_{1:T},\rnnparam^{(t)}) \times p(U_T^n = 1|\ts^n_{1:T},\rnnparam^{(t)})}{\sum_{U_T^n}p(\ts^n_{T+1:T+h}|U_T^n,\ts^n_{1:T},\rnnparam^{(t)}) \times p(U_T^n|\ts^n_{1:T},\rnnparam^{(t)})}
\end{align*}

where , $\mathbb{P}(U_T^n=1|\ts^n_{1:T},\rnnparam^{(t)})$ is the output of $\classifier$ at epoch $t$ and the predictive probability density $p(\ts^n_{T+1:T+h}|U_T^n=1,\ts^n_{1:T})$ is compute with the output of $\predictor$ at epoch $t$.

To compute our final loss function, we have to define the predictive probability density $p_\theta(\ts^n_{T+1:T+h}|U_T^n,\ts^n_{1:T})$.

In our model final equation~\ref{eq:full:model}, we assume that $\varepsilon^n_{T+i}$ are independent and follow a Laplace distribution.

We have for the regime 0 :

\begin{align*}
p_\theta(\ts^n_{T+1:T+h}|U_T^n=0,\ts^n_{1:T}) &= \prod_{i=1}^{h}\frac{1}{2b}e^{\frac{1}{b}|\ts^n_{T+i}  - \tspred^{stat,n}_{T+i|T}|} \\
& = \frac{1}{(2b)^h}e^{\frac{1}{b}\sum_{i=1}^{h}|\ts^n_{T+i}  - \tspred^{stat,n}_{T+i|T}|}
\end{align*}

and for the regime 1 :
\begin{align*}
p_\theta(\ts^n_{T+1:T+h}|U_T^n=0,\ts^n_{1:T}) &= \frac{1}{(2b)^h}e^{\frac{1}{b}\sum_{i=1}^{h}|\ts^n_{T+i}  - \tspred^{stat,n}_{T-w+i|T}-\meants^n_T \cdot \predictor(\fullconcatinput^n_T)_{T+i} |}
\end{align*}

with $b$ and constant that we fixed at $ \meants^n_T = \sum_{i = 1}^{w}  \ts^n_{T-w+i}/w$.

Finally, we have all the values to compute the loss function :

$$
\theta\mapsto - \omega_{\rnnparam^{(t)}}^n \log p_\theta(\ts^n_{T+1:T+h},U_T^n=1|\ts^n_{1:T})
- (1-\omega_{\rnnparam^{(t)}}^n) \log p_\theta(\ts^n_{T+1:T+h},U_T^n=0|\ts^n_{1:T})\,, \\
$$
where no gradient is computed on $\omega_{\rnnparam^{(t)}}^n$.
